# Data

> Quantities, characters, or symbols on which operations are performed by a 
> computer, considered collectively. Also (in non-technical contexts): 
> information in digital form.
>
> \- Oxford English Dictionary


<!--
Information becomes data when it takes on a structure.

some data has a structure that is inherent to the information
Data is actionable, either by humans or by computers by putting it in some sort 
of structure (including natural language: syntax and grammar), text isn't 
unstructured it just hasn't been restructured to make it actionable

two structures; inherent structure and structure we impose on the data
The process of (re)structuring data is at least as important as the analysis 
we do afterward.

Things inform your data structure: inherent structure in data, and the question
you want to ask?

Ex you might throw the structure away ex with a lot of LLMs through the grammatical
structure away.

You will almost always have multiple versions of data in different structures
so you can ask different questions

Note: there are other definitions of data structure from other disciplines
(ex. in comp sci), we will not be using those definitions
-->

If you are doing research, chances are you are going to be working with data
in some form or another. Even if you do not use the traditional rows and
columns people most often associate with "data", you certainly working with
information that has been translated into a form that is more easily shared
or stored for later use, aka data. Information becomes data when we give it some
sort of structure. That could be the previously mentioned rows and columns, or 
something more human readable, like the grammar and syntax of natural language.

One of the challenges of working with data is figuring out the best way, or at
least a good way, to store it. Thankfully, we have mostly left the 20th
century's favorite way of storing data: paper in filing cabinets. Our new
methods are definitely less flammable but but also can have a higher barrier to
entry. There are also a greater variety in types of data stores, which means you
need to make a decision on which one to use. Ultimately, the best way to store
your data will depend on the data you want to store, specifically the data type
and structure.

## Data Types

Your data's type tells the computer what sorts of operations make sense to do
with it. Common data types include integers, decimal or floating point numbers,
characters (text), Boolean values (TRUE and FALSE) and date/times. Data type
also influences how data is stored. Unlike other characteristics we will discuss later, an individual data point can have a type.

Data type is not necessarily inherent to a piece of information, though it is
often strongly suggested. For example, you can store the value of 1 as the
integer 1, the decimal 1.0, the text "1", the Boolean TRUE, or the date
1900-01-01 00:00:00. The underlying information is the same (1), but the
computer will treat each of these pieces of data differently. It will not allow
you to add "1" to another number, and it may have issues combining the integer
or decimal 1.0 with text. Text itself can only be stored as characters.

## Data Structures

> A data structure is a collection of data values, the relationships among them,
> and the functions or operations that can be applied to the data.
>
> Wikipedia

Unlike with data types, a single piece of data cannot have a data structure.
This is because data structure is about capturing relationships between data
points, there isn't any meaningful information to be gained from the
relationship of a data point with itself. We will discuss five of the most
common data: tabular/columnar, graph, tree, categorical/relational, and spatial
data. In practice, data often exhibits traits from multiple data structures. 

As mentioned previously, data is information translated into a form to store for
later use. The type of information you work with does not dictate the type of
data structure you must use. However, some types of information
lend themselves to particular data structures over others. When trying to decide
what data structure to use to represent information from your research, the key
is to focus on what sorts of questions you would like to ask. Each data
structure will make some types of questions easier to ask and others harder. If
you want to ask disparate types of questions, you may need to represent your
data using multiple structures. Because data structure has such
a large impact on how you interact with your data, the structure you choose is
just as important as the analysis you do afterward.

### Columnar/Tabular

<!--
Convention is rows are observations and columns are your variables,
value associated with a given variable for a given observation lives in that
cell

Discrete (brief) observations of a particular thing, 
dealing with a class of object where every observation of one of those objects
can be described using the same set of variables

sorting is easy

better at working across the entire data structure

location and scale aka spread
-->


```{r format_cars, echo=FALSE}

library(data.table)
library(stringr)

makers = c('chevrolet', 'honda', 'toyota', 'volkswagen', 'volvo')

car_origin = data.frame(origin=1:3, origin_region=c('USA', 'Europe', 'Japan'))

mpg_all = fread('data/autos_mpg.csv')

mpg = merge(mpg_all, car_origin, by='origin')

mpg$make = str_split_i(mpg$`car name`, ' ' , i = 1)
mpg$model = str_split_fixed(mpg$`car name`, ' ', 2)[, 2]

mpg = mpg[`model year`==78 & make %in% makers, ]

coi = c('make', 'model', 'mpg', 'cylinders', 'horsepower', 'weight',
        'origin_region')

mpg = mpg[, ..coi]

setnames(mpg, 'origin_region', 'origin')


```


Tabular or columnar data is made up of rows and columns, collectively referred
to as a table. Each row represents an observation, and each column a measurement
or aspect of that observation. In the table below, each row is an observation of
a car model released in 1978. 

```{r display_mpg, echo=FALSE}

mpg

```


### Graph

Graph data structure explicitly captures between entities. In graphs, 
relationships are defined by position and distance. Relationships can be 
directional 

### Tree 

Tree data is very similar to graph data in many ways. Trees also explicitly
capture relationships between entities. However, in trees the relationships
are hierarchical. The number of nodes separating an entity from a common root
defines that entity's place in the hierarchy. This is also 

### Categorical - Relational 

In categorical data, category membership is the organizing principle of the 
data structure. Entities can have membership in multiple categories, and
relationships are defined between categories instead of the entities themselves.
As such, relationships tend to represent complex processes rather than
individual entities.

### Spatial

In spatial data, the organizing principle is location. This can be geographic 
location (ex latitude/longitude) or relative location on an arbitrary grid
(ex. [2, 4])  

## Picking a Data Structure

As mentioned previously, data represents information from the real world. And
all information in the real world can be thought of in a larger context, like
zooming out with a camera lens. You can use this context to identify the best
structure to use for your data. If you view your data as observation of some
larger phenomenon, tabular data could be a good fit. If you are more concerned
about how various entities relate to each other in the context of your research,
a graph may be more appropriate.

Additionally, many research questions are complex enough that one data structure
on its own may not be sufficient to capture all the information necessary to
answer the question. It it is very common to combine data structures to capture
more information that one could on its own. Most observation based data sets are
actually a combination of tabular data and categorical - relational data, which
captures information about the relationships between observations. Phylogenetic
trees, the data structure most commonly used in the study of evolution, is a
combination of the graph and tree data structures.




